---
title: "Constrained analyses applied to an interaction network"
format: 
  html:
    toc: true
    toc-depth: 3
editor: source
author: 
  - name: "Lisa Nicvert"
    orcid: 0009-0006-5763-0865
  - name: "Hervé Fritz"
    orcid: 0000-0002-7106-3661
  - name: "Stéphane Dray"
    orcid: 0000-0003-0153-1105
embed-resources: true
params:
  thr: 1
  colco: "darkred"
  colli: "cornflowerblue"
  dataset: "Peru1"
bibliography: references.bib
---

```{r}
#| code-fold: true
#| output: false

library(here)

library(ggplot2)
library(patchwork)

library(ade4)
library(adegraphics)

devtools::load_all()

# Paths
figures_folder <- here("figures")
out_folder <- here("outputs", "02_example_network", params$dataset)
input_folder <- here("outputs", "01_clean_data", params$dataset)

set.seed(42)
```

## Introduction

In this document, we analyze dataset `r params$dataset` from ANDEAN frugivory dataset [@dehlinga] with correspondence analyses methods.

## Read data

First, read formatted data.

```{r}
interactions <- read.csv(file.path(input_folder, "interactions.csv"))
plant_traits <- read.csv(file.path(input_folder, "plant_traits.csv"))
bird_traits <- read.csv(file.path(input_folder, "bird_traits.csv"))
```

## Format data

We remove some redundant traits (`BillLength` and `meanL`).

```{r}
# Remove "useless" traits
# bird_traits <- bird_traits |>
#   dplyr::select(-BillLength)
# 
# plant_traits <- plant_traits |>
#   dplyr::select(-meanL)
```

Then, we remove the species that have `NA` traits.

```{r filter_out_incomplete_traits}
bird_traits <- bird_traits |>  
  tidyr::drop_na()

plant_traits <- plant_traits |> 
  tidyr::drop_na()

# Keep the table with traits for export
bird_traits_codes <- bird_traits
plant_traits_codes <- plant_traits
```

We format the interaction matrix and filter out species that interact with $\leq$ `r params$thr` partners.

```{r filter_data}
# Convert data.frame to matrix
interactions <- df_to_matrix(interactions,
                             rows = "plant_species_code",
                             columns = "bird_species_code")

# Remove species that interact with few partners
interactions <- filter_matrix(interactions, thr = params$thr)

# Format data
plant_traits <- plant_traits |> 
  tibble::column_to_rownames("plant_species_code") |> 
  dplyr::select(-plant_species)

bird_traits <- bird_traits |> 
  tibble::column_to_rownames("bird_species_code") |> 
  dplyr::select(-bird_species)
```

```{r}
# Filter
p <- intersect(rownames(interactions), 
               rownames(plant_traits))
b <- intersect(colnames(interactions), 
               rownames(bird_traits))

plant_traits <- plant_traits[p, ]
bird_traits <- bird_traits[b, ]
interactions <- interactions[p, b]
```

```{r}
# Rename columns
plant_traits_labels <- c("fruit length", "fruit diameter", "plant height", "crop mass")
names(plant_traits_labels) <- colnames(plant_traits)

bird_traits_labels <- c("Kipp's index", "bill length", "bill width", "body mass")
names(bird_traits_labels) <- colnames(bird_traits)
```

```{r}
# Get data dimensions
r <- nrow(interactions)
c <- ncol(interactions)
l <- ncol(plant_traits)
k <- ncol(bird_traits)
```

```{r}
# Create viridis palette for rows
colrow <- viridisLite::viridis(nrow(interactions), 
                               option = "G", 
                               end = 0.9)
# Create viridis palette for columns
colco <- viridisLite::viridis(ncol(interactions), 
                              option = "F", 
                              end = 0.9)
```


## CA

First, perform simple correspondence analysis (CA): this will be useful for variance partitioning and for computing CCA and dc-CA with ade4.

```{r}
ca <- dudi.coa(interactions, 
               scannf = FALSE, 
               nf = min(c(r - 1, c - 1)))
```

## CCA with plant traits

Then, perform constrained correspondence analysis (CCA) constrain with plant traits.

```{r}
cca_plants <- pcaiv(dudi = ca, 
                    df = plant_traits,
                    scannf = FALSE,
                    nf = min(c(r - 1, c, l)))
```

```{r, fig.height=4, fig.width=6}
#| code-fold: true

plot_eig(sqrt(cca_plants$eig), showrank = TRUE) + 
  ggtitle("Square-root of the eigenvalues of the CCA (correlations)")

plot_eig(cca_plants$eig/sum(ca$eig), showrank = TRUE) + 
  ggtitle("Part of explained variance for each axis of the CCA")
```

Correlations are decent (0.3-0.5), but the part of explained variation for each axis is very low.


### Correlation circle

```{r}
#| code-fold: true

# We invert second axis for easier interpretation (top = larger trait values)
cor <- cca_plants$cor
# cor$RS2 <- -cor$RS2
plot_corcircle(cor = cor, 
               col = params$colli,
               label = plant_traits_labels,
               eig = cca_plants$eig)

ggsave(file.path(figures_folder, "Fig_C1a_cca_plants_corcircle.png"),
       width = 16, height = 13, units = "cm")
```

### Biplots

Let's plot birds WA scores ($V^\star$) (`co`) at the mean of plants LC (linear combinations) scores $Z$ (`l1`) (bird niches).

```{r}
#| code-fold: true

# Plot bird niches
# Second axis is inverted
l1 <- cca_plants$l1
# l1[, 2] <- - l1[, 2]

co <- cca_plants$co
# co[, 2] <- - co[, 2]


multiplot(indiv_row = l1,
          indiv_col = co, 
          eig = cca_plants$eig, 
          row_color = params$colli, 
          col_color = params$colco, 
          xlim = c(-6, 6),
          max.overlaps = 15)
ggsave(file.path(figures_folder, "Fig_C1b_cca_plants_bniches.png"),
       width = 13, height = 13, units = "cm")
```
We could also represent plants niches with plants WA (weighted averaging) scores $U^\star$ (`ls`) at the mean of birds scores $V$ (`c1`).

## CCA with bird traits

Then, we constrain axes with bird traits.

```{r}
ca_t <- dudi.coa(t(interactions), 
                 scannf = FALSE, 
                 nf = min(c(r - 1, c - 1)))

cca_birds <- pcaiv(dudi = ca_t, 
                   df = bird_traits,
                   scannf = FALSE,
                   nf = min(c(c - 1, r, k)))
```


```{r, fig.height=4, fig.width=6}
#| code-fold: true

plot_eig(sqrt(cca_birds$eig)) + ggtitle("Square-root of the eigenvalues of the CCA (correlations)")

plot_eig(cca_birds$eig/sum(ca$eig)) + 
  ggtitle("Part of explained variance for each axis of the CCA")
```

### Correlation circle

```{r}
#| code-fold: true

cor <- cca_birds$cor
cor[, 1]  <- -cor[, 1]
plot_corcircle(cor = cor, 
               col = params$colco,
               label = bird_traits_labels,
               eig = cca_birds$eig)

ggsave(file.path(figures_folder, "Fig_D1a_cca_birds_corcircle.png"),
       width = 14, height = 11, units = "cm")
```

### Biplots

We plot plants WA scores ($V^\star$) (`co`) at the mean of birds LC scores $Z$ (`l1`) (plants niches).

```{r}
#| code-fold: true

# Invert first axis
l1 <- cca_birds$l1
l1[, 1] <- - l1[, 1]

co <- cca_birds$co
co[, 1] <- - co[, 1]

# Plants niches
multiplot(indiv_row = l1,
          indiv_col = co, # WA cols = plants
          eig = cca_birds$eig, 
          row_color = params$colco, 
          col_color = params$colli, 
          # xlim = c(-2, 7),
          # ylim = c(-6, 5),
          max.overlaps = 15) +
  ggtitle("Plant niches")
ggsave(file.path(figures_folder, "Fig_D1b_cca_birds_pniches.png"),
       width = 13, height = 11, units = "cm")
```
## dc-CA

Now we perform a dc-CA, constrained by plant and bird species traits.

```{r}
dcca <- dpcaiv(dudi = ca, 
               dfR = plant_traits,
               dfQ = bird_traits,
               scannf = FALSE, 
               nf = min(k, l))
```

```{r, fig.height=4, fig.width=6}
#| code-fold: true

plot_eig(sqrt(dcca$eig)) + ggtitle("Square-roots of eigenvalues of the dc-CA")

plot_eig(dcca$eig/sum(ca$eig)) + 
  ggtitle("Part of explained variance for each axis of the dc-CA")
```

### Correlation circle

```{r corcircle}
#| code-fold: true

corR <- dcca$corR
# corR[, 2] <- -corR[, 2] 

corQ <- dcca$corQ
# corQ[, 2] <- -corQ[, 2] 
plot_corcircle(cor = corR, 
               col = params$colli,
               label = plant_traits_labels,
               cor2 = corQ,
               col2 = params$colco,
               lty = "longdash",
               lty2 = "solid",
               label2 = bird_traits_labels,
               eig = cca_plants$eig)
ggsave(file.path(figures_folder, "Fig_3a_dcca_corcircle.png"),
       width = 15, height = 11, units = "cm")
```

### Biplots

We plot the biplots with a scaling appropriate to represent species niches:

-   to represent plants niches: plants WA scores $U^\star$ (`lsR`) at the mean of birds LC scores $V$ (`c1`).
-   to represent bird niches: birds WA scores ($V^\star$) (`lsQ`) at the mean of plants LC scores $U$ (`l1`).

```{r}
#| code-fold: true

lsR <- dcca$lsR
# lsR[, 2] <- - lsR[, 2]

c1 <- dcca$c1
# c1[, 2] <- - c1[, 2]

l1 <- dcca$l1
# l1[, 2] <- - l1[, 2]

lsQ <- dcca$lsQ
# lsQ[, 2] <- - lsQ[, 2]

# Thr 1
# xlim <- c(-4, 7)
# ylim <- c(-4, 7)

# Thr 2
xlim <- c(-4, 9)
ylim <- c(-11, 4)
grad <- 4

# Plant niches
multiplot(indiv_row = lsR,
          indiv_col = c1, 
          eig = dcca$eig, 
          row_color = params$colli, 
          col_color = params$colco, 
          xlim = xlim,
          ylim = ylim,
          grad = grad,
          max.overlaps = 15)
ggsave(file.path(figures_folder, "Fig_3b_dcca_pniches.png"),
       width = 11, height = 11, units = "cm")

# Bird niches
multiplot(indiv_row = l1,
          indiv_col = lsQ,
          eig = dcca$eig, 
          row_color = params$colli, 
          col_color = params$colco,
          xlim = xlim,
          ylim = ylim,
          grad = grad,
          max.overlaps = 15)
ggsave(file.path(figures_folder, "Fig_3c_dcca_bniches.png"),
       width = 11, height = 11, units = "cm")
```

## Variance decomposition

The analyses above are linked through the following variance decomposition:

$$\underbrace{I(\text{total})}_{[t]} = \underbrace{I(\text{birds traits})}_{[c]} + \underbrace{I(\text{plants traits})}_{[a]} + \underbrace{I(\text{matching})}_{[b]} + \underbrace{I(\text{residual})}_{[d]}$$
where $I(x)$ denotes the inertia of $x$ (also denoted by letters between brackets).

### Figure

```{r}
#| code-fold: true

# Graph parameters
xmax <- 15 
ymax <- 10
width <- 5 # width of resource/consumer inertia
rpad <- 0.1 # padding for outline of the residual inertia
mar <- 0.1
inline <- 0.04

# The rectangles
df_rect <- data.frame(xmin = c(0, width+inline, 0+inline, 0+inline, width+rpad),
                      xmax = c(xmax, xmax-inline, width, width, xmax-rpad), 
                      ymin = c(0, 0+inline, width+inline, 0+inline, width+rpad), 
                      ymax = c(ymax, width, ymax-inline, width, ymax-rpad),
                      group = c("tot", 
                                "consumer_traits", "resource_traits",  
                                "matching", "res"))

# Text annotations
annotate <- data.frame(x = c(xmax, 
                             width+(xmax-width)/2, width/2, 
                             width/2, width+(xmax-width)/2),
                       y = c(ymax+mar, 
                             width/2, width+(ymax-width)/2, 
                             width/2, width+(ymax-width)/2),
                       label = c("total inertia [t]", 
                                 "consumer traits only [c]", "resource traits only [a]", 
                                 "matching [b]", "residual inertia [d]"),
                       group = df_rect$group)

palette_fill <- c("white", "darkred", "cornflowerblue", "darkorchid", "transparent")
names(palette_fill) <- df_rect$group

palette_col <- c("black", "#470000", "#1B61E4", "darkorchid4", "grey40")
names(palette_col) <- annotate$group

ggplot(df_rect, aes(col = group)) +
  geom_rect(aes(xmin = xmin, xmax = xmax,
                ymin = ymin, ymax = ymax,
                group = group, fill = group),
            alpha = c(1, rep(0.7, 3), 0),
            linetype = c(rep("solid", 4), "dashed"),
            show.legend = FALSE) +
  geom_label(data = annotate,
            aes(x = x, y = y, label = label),
            label.size = 0.5,
            vjust = c(0, 0.5, 0.5, 0.5, 0.5), 
            hjust = c(1, 0.5, 0.5, 0.5, 0.5),
            size = 4,
            show.legend = FALSE) +
  coord_fixed() +
  ylim(0, ymax + 0.5) +
  scale_fill_manual(values = palette_fill) +
  scale_color_manual(values = palette_col) +
  ggplot2::theme_void()

ggsave(file.path(figures_folder,
                 "Fig_2_variance_partitioning.png"),
       width = 15, height = 10.5, units = "cm")
```

### Values

Here, we correct the variance decomposition using the permutational form of adjustment from @peres-neto2006 using the `ade4::dvaripart` function.

```{r}
vp <- dvaripart(Y = ca,
                dfR = plant_traits, 
                dfQ = bird_traits,
                nrepet = 999)

vp
```

## Species details tables (Appendix A)

The code below exports trait tables with species names and codes.

```{r}
# Birds
bird_traits_codes$bird_species <- gsub("_", " ",
                                       bird_traits_codes$bird_species)
bird_traits_codes$bird_species <- paste0("\\textit{",
                                         bird_traits_codes$bird_species,"}")

if (!dir.exists(out_folder)) {
  dir.create(out_folder, recursive = TRUE, showWarnings = FALSE)
}

write.table(bird_traits_codes, sep = " & ", 
            eol = " \\\\\n",
            quote = FALSE, 
            col.names = FALSE, row.names = FALSE, 
            file = file.path(out_folder, "birds_latex.txt"))

# Plants
plant_traits_codes$plant_species <- gsub("_", " ",
                                         plant_traits_codes$plant_species)
plant_traits_codes$plant_species <- paste0("\\textit{",
                                             plant_traits_codes$plant_species,"}")

write.table(plant_traits_codes, sep = " & ", 
            eol = " \\\\\n",
            quote = FALSE, 
            col.names = FALSE, row.names = FALSE, 
            file = file.path(out_folder, "plants_latex.txt"))
```

## Traits distribution (Appendix B)

Plot histograms of trait values.

```{r, fig.width=8, fig.height=3}
#| code-fold: true

plant_traits_labels_units <- paste(plant_traits_labels, 
                                   c("(mm)", "(mm)", "(m)", "(g)"))
names(plant_traits_labels_units) <- names(plant_traits)

glist <- list()
for (colname in colnames(plant_traits)) {
  gr <- ggplot(plant_traits) +
    geom_histogram(aes(x = .data[[colname]]),
                 fill = params$colli,
                 col =  params$colli,
                 alpha = 0.8) +
    annotate("segment", 
             x = median(plant_traits[[colname]]),
             xend = median(plant_traits[[colname]]),
             y = 0, yend = Inf,
             linetype = "dashed") +
    xlab(plant_traits_labels_units[colname]) +
    ylab("species count") +
    theme_linedraw()
  
  glist <- c(glist, list(gr))
}

wrap_plots(glist) +  
  plot_layout(axis_titles = "collect_y", nrow = 1)

ggsave(file.path(figures_folder, "Fig_B1a_plant_traits.png"),
       width = 26, height = 7, units = "cm")
```

```{r, fig.width=8, fig.height=3}
#| code-fold: true

bird_traits_labels_units <- paste(bird_traits_labels, 
                                   c("", "(mm)", "(mm)", "(g)"))
names(bird_traits_labels_units) <- names(bird_traits_labels)

glist <- list()

for (colname in colnames(bird_traits)) {
  g <- ggplot(bird_traits) +
    geom_histogram(aes(x = .data[[colname]]),
                   fill = params$colco,
                   col =  params$colco,
                   alpha = 0.8) +
    annotate("segment", 
             x = median(bird_traits[[colname]]),
             xend = median(bird_traits[[colname]]),
             y = 0, yend = Inf,
             linetype = "dashed") +
    xlab(bird_traits_labels_units[colname]) +
    ylab("species count") +
    theme_linedraw()
  
  glist <- c(glist, list(g))
}

wrap_plots(glist) +  
  plot_layout(axis_titles = "collect_y", nrow = 1)

ggsave(file.path(figures_folder, "Fig_B1b_bird_traits.png"),
       width = 26, height = 7, units = "cm")
```

## Residuals (Appendix E)

For constrained analyses, we have two sets of scores for constrained species: comparing them is a way to measure the model fit.

### CCA with plant traits

Let's plot plant coordinates:

- from their traits: the LC scores $Z_0$ (non-normalized) (`li`) 
- from birds' scores ("latent traits"): the WA scores $U^\star$ (`ls`)

```{r}
#| code-fold: true

ls <- cca_plants$ls
# ls[, 2] <- - ls[, 2]

li <- cca_plants$li
# li[, 2] <- - li[, 2]

# ls -> li
s.match(ls, li, xax = 1, yax = 2, 
        plabels.col = params$colli, 
        plabels.cex = 0.6)
```

### CCA with bird traits

Let's plot bird coordinates:

- from their traits: the LC scores $Z_0$ (non-normalized) (`li`) 
- from plants' scores ("latent traits"): the WA scores $U^\star$ (`ls`)

```{r}
#| code-fold: true

ls <- cca_birds$ls
ls[, 2] <- - ls[, 2]

li <- cca_birds$li
li[, 2] <- - li[, 2]

# ls -> li
s.match(ls, li, xax = 1, yax = 2, 
        plabels.col = params$colco,
        plabels.cex = 0.8)
```


### dcCA

Now, let's plot bird and plants coordinates:

- from their traits: the LC scores $U$ and $V$ (non-normalized) (respectively `li` and `co`) 
- from their "latent traits: the WA scores $U^\star$ and $V^\star$ (respectively `lsR` and `lsQ`)

```{r}
#| code-fold: true

li <- dcca$li
li[, 2] <- - li[, 2]

co <- dcca$co
co[, 2] <- - co[, 2]

percent <- (dcca$eig/sum(ca$eig))*100
xlab <- paste0("Axis 1 (", round(percent[1], 2), "%)")
ylab <- paste0("Axis 2 (", round(percent[2], 2), "%)")

axis_title_padding <- 1.3
padding <- 1

# Difference between niche and traits for plants
# lsR -> li
# png(file.path(figures_folder, "Fig_Eb_diff_plants.png"),
#     width = 15, height = 15, units = "cm", res = 300)
s.match(lsR, li, xax = 1, yax = 2, 
        plabels.col = params$colli,
        plabels.cex = 0.8,
        xlab = xlab, ylab = ylab,
        layout.heights = list(axis.xlab.padding = axis_title_padding,
                              bottom.padding = padding,
                              top.padding = padding),
        layout.widths = list(ylab.axis.padding = axis_title_padding,
                             left.padding = padding,
                             right.padding = padding))
# dev.off()

# Difference between niche and traits for birds
# lsQ -> li
# png(file.path(figures_folder, "Fig_Ea_diff_birds.png"),
#     width = 15, height = 15, units = "cm", res = 300)
s.match(lsQ, co, xax = 1, yax = 2, 
        plabels.col = params$colco,
        plabels.cex = 0.8,
        layout.heights = list(axis.xlab.padding = axis_title_padding,
                              bottom.padding = padding,
                              top.padding = padding),
        layout.widths = list(ylab.axis.padding = axis_title_padding,
                             left.padding = padding,
                             right.padding = padding))
# dev.off()
```

