---
title: "Analyze networks"
format: 
  html:
    toc: true
    toc-depth: 3
editor: source
author: 
  - name: "Lisa Nicvert"
    orcid: 0009-0006-5763-0865
embed-resources: true
code-fold: true
params:
  colco: "darkred"
  colli: "cornflowerblue"
bibliography: references.bib
---


```{r, message = FALSE, warning= FALSE}
library(here)

library(ggplot2)
library(patchwork)

library(ade4)
library(adegraphics)

devtools::load_all()
```

```{r}
thr <- 2

nrepet <- 999

figures_folder <- here("figures")
input_folder <- file.path(here("outputs/01_clean_data"))
```

## Read data

```{r read_data}
dirs <- list.dirs(input_folder, full.names = FALSE)
dirs <- dirs[dirs != ""]

inter_list <- list()
pt_list <- list()
bt_list <- list()
for (dir in dirs) {
  inter <- read.csv(file.path(input_folder, dir, 
                              "interactions.csv"))
  pt <- read.csv(file.path(input_folder, dir, 
                              "plant_traits.csv"))
  bt <- read.csv(file.path(input_folder, dir, 
                              "bird_traits.csv"))
  
  inter_list[[dir]] <- inter
  pt_list[[dir]] <- pt
  bt_list[[dir]] <- bt
  }
```

```{r}
pnames_list <- list()
bnames_list <- list()
for (i in 1:length(inter_list)) {
  # Filter matrix
  intermat <- df_to_matrix(inter_list[[i]], 
                           rows = "plant_species_code",
                           columns = "bird_species_code")
  intermat <- filter_matrix(intermat, thr = thr)
  
  # Prepare traits
  pnames <- pt_list[[i]] |> 
    dplyr::select(plant_species_code, plant_species)
  pt_filtered <- pt_list[[i]] |>
    tibble::column_to_rownames("plant_species_code") |> 
    tidyr::drop_na() |> 
    dplyr::select(-plant_species) 
  
  bnames <- bt_list[[i]] |> 
    dplyr::select(bird_species_code, bird_species)
  bt_filtered <- bt_list[[i]] |>
    tibble::column_to_rownames("bird_species_code") |> 
    tidyr::drop_na() |> 
    dplyr::select(-bird_species) 
  
  # Filter
  p <- intersect(rownames(intermat), rownames(pt_filtered))
  b <- intersect(colnames(intermat), rownames(bt_filtered))
  
  pt_filtered <- pt_filtered[p, ]
  bt_filtered <- bt_filtered[b, ]
  intermat <- intermat[p, b]
  
  # Store names
  dataname <- names(inter_list)[i]
  pnames_list[[dataname]] <- pnames
  bnames_list[[dataname]] <- bnames
  
  # Update lists
  inter_list[[i]] <- data.frame(intermat)
  pt_list[[i]] <- pt_filtered
  bt_list[[i]] <- bt_filtered
}
```

## Correspondence analyses

```{r}
Rcor_df <- data.frame()

for (i in 1:length(inter_list)) {
  inter <- inter_list[[i]]
  pt <- pt_list[[i]]
  bt <- bt_list[[i]]
  
  r <- nrow(inter)
  c <- ncol(inter)
  l <- ncol(pt)
  k <- ncol(bt)
  
  # CA
  ca <- dudi.coa(inter, scannf = FALSE,
                 min(c(r - 1, c - 1)))
  
  vp <- dvaripart(Y = ca, 
                  dfR = pt, dfQ = bt)
  
  R2 <- vp$R2.adj
  
  pval <- vp$test$pvalue
  

  Rcor <- data.frame(dataset = rep(names(inter_list)[i], 4), 
                     type = c("residual", 
                              "bird_traits", "plant_traits",
                              "matching"),
                     R2 = c(R2["d"], 
                            R2["c"], R2["a"], 
                            R2["b"]),
                     p = c(NA, pval[2], pval[1],
                           max(pval[c(3, 4)])))
  
  Rcor_df <- rbind(Rcor_df, Rcor)
}
```

## Results analysis

First, check that the $R^2$ sum to one.

```{r}
Rcor_df |> 
  dplyr::group_by(dataset) |> 
  dplyr::summarize(sum = sum(R2))

# Ecuador 3 is of dimension 3x5, so there are too many traits and R2 = 1
# we remove it
Rcor_df <- Rcor_df |> 
  dplyr::filter(dataset != "ANDEAN_Ecuador3")
```

```{r}
# Prepare df
Rcor_df <- Rcor_df |> dplyr::mutate(signif = p > 0.05)

# Add columns
Rcor_df <- Rcor_df |> 
  dplyr::mutate(dataset_name = gsub("^ANDEAN_", "", dataset), .after = dataset) |> 
  dplyr::mutate(dataset_name = gsub("_", " ", dataset_name)) |> 
  dplyr::mutate(type_name = gsub("_", " ", type), .after = type)

# Reorder factors
Rcor_df$type <- factor(Rcor_df$type, 
                       levels = c("residual", "bird_traits", "plant_traits", "matching"))
Rcor_df$type_name <- factor(Rcor_df$type_name, 
                            levels = c("residual", "bird traits", "plant traits", "matching"))
```


Below, we summarize the results by dataset:

```{r}
coltype <- viridisLite::viridis(n = length(levels(Rcor_df$type_name)), 
                                begin = 0.1, end = 0.9)
names(coltype) <- levels(Rcor_df$type_name)

# Dataset order
subset <- Rcor_df[Rcor_df$type == "residual", ]
ord <- order(subset$R2[subset$type == "residual"])
dataset_order <- subset$dataset_name[rev(ord)]

Rcor_df_bar <- Rcor_df
Rcor_df_bar$dataset_name <- factor(Rcor_df_bar$dataset_name, 
                                   levels = dataset_order, ordered = TRUE)
```


```{r}
(gbar <- ggplot(Rcor_df_bar,
                aes(y = dataset_name, x = R2, fill = type_name)) +
  geom_col() +
  geom_vline(xintercept = c(0, 1), linetype = "dashed") +
  scale_fill_manual(values = coltype, name = "Origin of variation") +
  guides(fill = guide_legend(reverse = TRUE)) +
  ylab("Dataset") +
  theme_minimal() +
  theme(legend.position = "bottom"))
```


```{r}
(gbox <- ggplot(Rcor_df,
               aes(y = forcats::fct_rev(type_name), 
                   x = R2, color = type_name)) +
  geom_boxplot(show.legend = FALSE) +
  geom_jitter(width = 0, height = 0.2,
              show.legend = FALSE) +
  stat_summary(fun = "median", 
               geom = "label", 
               aes(label = round(after_stat(x), 3)),
               size = 3,
               show.legend = FALSE) +
  scale_color_manual(values = coltype) +
  ylab("Origin of variation") +
  theme_minimal())
```

```{r}
(gbar + gbox) + 
  plot_layout(guides = 'collect',
              axis_titles = "collect") &
  theme(legend.position='bottom'
        # axis.text.y = element_text(angle = 90, vjust = 0.5)
        ) &
  scale_x_continuous(breaks = seq(0, 1, by = 0.2), 
                     limits = c(-0.1, 1.1)) &
  xlab(bquote("Corrected"~R^2))

ggsave(file.path(figures_folder, "Fig4_dcca_var_networks.png"),
       width = 18, height = 10, units = "cm")
```


This graph shows that for most datasets, the largest part of variation is residual and cannot be attributed to the traits considered here (listed below):

- plants: fruit length and diameter, plant height and crop mass;
- birds: Kipp's index (wing pointedness), bill length and width and body mass.

```{r}
unique(lapply(pt_list, names))

unique(lapply(bt_list, names))
```

Moreover, we can see that for some datasets, negative parts of variations for plant traits of matching make the result more difficult to interpret (basically, it means that for these datasets, matching cannot be reliably assessed.)

```{r}
neg <- Rcor_df |> 
  dplyr::group_by(dataset) |> 
  dplyr::summarize(has_neg = any(R2 < 0)) |> 
  dplyr::filter(has_neg) 

paste(nrow(neg), "datasets have negative R2 values:", 
      paste(neg$dataset, collapse = ", "), collapse = " ")
```

```{r}
Rcor_df |> 
  dplyr::arrange(dataset, type) |> 
  DT::datatable()
```


The table below summarizes these statistics.

```{r}
Rcor_df |> 
  dplyr::group_by(type) |> 
  dplyr::summarize(mean_R2 = mean(R2),
                   median_R2 = median(R2),
                   sd_R2 = sd(R2),
                   prop_signif = sum(signif)/length(unique(dataset))) |> 
  DT::datatable()
```
